\chapter{Literature Review}

\section{Introduction}

The purpose of a literature review is to identify existing areas of research relating to the given topic, alongside the any areas that are lacking in
research literature. It also serves to provide an understanding of the topics that are intended to be discussed and expanded upon during the project.

\section{Review of existing literature}

Scientific literature regarding any subset of physics is traditionally written for those with a strong background of 
physics, mathematics, or research. Because of this, people outside of such disciplines, such as those in computer
science including the game development fields, may be discouraged to use these resources due to the formatting,
terminology or mathematic symbology that may unfamiliar to them. The result is developers who miss out on 
learning new techniques, algorithms, or optimisation that may benefit their software or game. This literature 
review, and the following report evaluates foundational physics papers and their accessibility to those not 
discipined in physics or mathematics through the creation of a particle physics simulation in C++. 

This review aims to source research within particle physics, and computer science.

\subsection{Foundational Physics}

This first category of literature consists of foundational physics papers, such as \bookcite{standard-model-workbook}. While reports such as these serve to
explain and give context to the bedrock of particle physics, they do so whilst presenting a developer experience barrier due to how its content is often dense mathematics.

From a software engineering perspective, the primary concern would be the use of abstract notation which, while logically might be familiar, are syntaxtically unorthodox.
An example of this is $\sum$ which represents iterative accumulation, comparable that to a programmatical \textit{for loop}. Because of this notation, software
and game developers would first need to translate the content from the paper into pseudo code.

\subsection{Computational Physics}

Computational literature such as \bookcite{steinhauserComputerSimulationPhysics2012} are often based on the implementation of the 
theory established by foundational literature. Because of this, it is less mathematically dense, and more focal on the practical
implementation of the functions and concepts introduced.

However, with this particular example, the argument could be made that it has the inverse problem than the foundational paper mentioned above,
focusing heavily on the basics of computer science, such as explaining binary and arithmetic-in-code, rather than the actual implementation of the mathematics.

\subsection{Developer Experience \& Accessibility}

The disparity between these two types of academic literature creates a gap between the the Developer Experience (\textbf{DX}), which can
result in longer development time as considerable time may be required to 'reverse-engineer' the mathematics.

Furthermore, it should be noted that aims of the reseacher and developer may differ.

While the researcher may be showing the metholodgy used to achieve a specified result. The developer often times would also
also aim for efficiency and scalablity in the algorithms used; something that the researcher would not have considered.

\subsection{Selected Studies on Particle Physics Simulation}

The following subsections present a comparison of a series of papers that relate to particle physics simulation, and, where possible,
address how accessible their content are for non-specialists.

\subsubsection*{Computer simulation in data analysis: A case study from particle physics}

In the paper \bookcite{falkenburgComputerSimulationData2024} a framework for processing data from large particle colliders is presented.
The paper includes key terms and notation such as:
\begin{enumerate}
    \item Likelihood function: $\mathcal{L}(\theta)=\prod_u P(x_i|\theta)$
    \item Bayesian prior: $\pi(\theta)$
    \item Markov Chain Monte Carlo: $\alpha=\min\!\left(1,\frac{\pi(\theta')}{\pi(\theta)}\right)$
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Lexical}: `Likehood' and `Prior' terms are only defined in footnotes, which may go unnoticed
    \item \emph{Notational}: Product symbol and superscript notation assume that the reader is familiar with certain mathematical notation and probablity theory.
\end{itemize}

This paper arguably demonstrates that well-structured simulation papers can still be inacessible to developers.

\subsubsection*{Differentiable simulations for particle tracking in accelerators: Analysis, benchmarking, and optimization}

In the paper \bookcite{huhnDifferentiableSimulationsParticle2025}{} a framework is introduced to allow for gradient-based optimisation of particle trajectories.
The paper includes key terms and notation such as:

\begin{enumerate}
    \item Hamiltonian $H(\mathbf{z})$, a function of phase-space vector $\mathbf{z}$.
    \item Symplectic integrator
    \item Gradient of the loss function $\nabla_{\theta}L(\theta)$
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Notational}: Heavy use of matrix algebra and Poisson brackets
    \item \emph{Conceptual}: Phase-space dynamics may be unfamiliar to computer science and game programmers.
\end{itemize}

\subsubsection*{Particle simulation study on the transport of laser-accelerated proton beams in high-density plasmas}

In the paper \bookcite{xuParticleSimulationStudy2024}{}, a model is presented showing proton acceleration in dense plasmas using particle-in-cell (PIC) techniques.
The paper includes key terms and notation such as:

\begin{enumerate}
    \item Electric field $E = -\nabla\Phi$
    \item Lorenz force $\mathbf{F}=q(\mathbf{E}+\mathbf{v}\times\mathbf{B})$
    \item Normalised units (e.g., $a_0$, $\omega_{pe}$)
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Notational}: Cross product notation and normalised units
    \item \emph{Conceptual}: Plasma frequency $\omega_{pe}$ and normalisation may be unfamiliar in a physics context
\end{itemize}

\subsubsection*{Simulation Study on Closed-orbit Correction for Storage Ring of Wuhan Advanced Light Source}

The paper \bookcite{simulationStudyClosedorbit2026}{} presents an algorithm to correct closed-orbit correction for the storage ring of the Wuhan Advanced Light Source, focusing on beam stability.
The paper includes key terms and notation such as:

\begin{enumerate}
    \item Closed-orbit deviation $\Delta x(s)$
    \item Orbit response matrix $R_{ij}$
    \item Singular value decomposition (SVD)
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Notational}: Use of matrix calculus and SVD may be unfamiliar to some game developers
    \item \emph{Conceptual}: The idea of a `closed orbit' and its physical significance may not be initially known
\end{itemize}

\subsubsection*{Quantum simulation of 2D topological physics in a 1D array of optical cavities}

The paper \bookcite{luoQuantumSimulation2D2015}{} demonstrates a quantum simulation of 2D topological phases using a 1-D array of coupled optical cavities.
The paper includes key terms and notation such as:

\begin{enumerate}
    \item Tight-binding Hamiltonian $H = -J\sum_{\langle i, j\rangle} (a_i^\dagger a_j + \text{h.c})$
    \item Berry curvature $\Omega(k)$
    \item Chern number $C = \frac{1}{2\pi}\int_{\text{BZ}}\Omega(k)\,dk$.
    \item Pauli matrices $\sigma_x,\sigma_y,\sigma_z$
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Notational}: Heavy used of second-quantised operators, integrals, and complex matrix notation.
    \item \emph{Conceptual}: Topics such as the Chern number and Berry curvature are not commonly studied topics in computer science or game development
\end{itemize}

\subsubsection*{Performance Analysis of C and Python Parallel Implementations on a Multicore System Using Particle Simulation}

The paper \bookcite{asaduzzamanPerformanceAnalysisPython2024}{} compares C++ and Python implementation of a particle simulation on a multicore system, and focuses on optimisation,
memory usage, and scalability.
The paper includes key terms and notation such as:

\begin{enumerate}
    \item OpenMP pragmas (e.g `\#pragma omp paralell for')
    \item GIL (Global Interpreter lock)
    \item Speed-up factor $S = T_{\text{serial}}/T_{\text{parallel}}$.
    \item Amdahl's Law: $S_{\text{max}} = 1/(1-P + P/N)$.
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Lexical}: Terms such as `GIL' and `OpenML' are specifgic to HPC and may be unknown.
    \item \emph{Notational}: Speed-up formulae involve fractions and variables that may be unfamiliar.
\end{itemize}

\subsubsection*{Simulation of matter–antimatter creation on quantum platforms}

The paper \bookcite{burrelloSimulationMatterAntimatter2025}{} proposes a quantum algorithm to simulate matter-antimatter pair creation
The paper includes key terms and notation such as:

\begin{enumerate}
    \item Qubit state $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$.
    \item Controlled-NOT gate
    \item Trotterisation
    \item Quantum volume $V_Q$
\end{enumerate}

\textbf{Accessibility}

\begin{itemize}
    \item \emph{Notational}: Dirac notation, gate symbols, and Trotterisation forumula may be unfamiliar to some game developers. However, the those with a computer science background are more likely to understand gate symbols
    \item \emph{Conceptual}: The physical process of pair creation and its integration onto a quantum circuit may require background knowledge in high energy physics.
\end{itemize}

\subsection{Summary}

The selected studies indicate the particle-physics simulation papers usually contain notation and terminonaly that can be considered verbose or inaccessible to devlopers with a background in
physics, or mathematics.

None of these papersd provide an audit of these notational barriers, nor do they link such barriers to an actual effort at implementation.

And as such, this review has provided an insight to the current landscape of particle physics simulation literature, with identified key accessibility issues, and a potentionally justified need for
an audit, or key of notation and terminology used.

The subsequent chapters will present the methodology used to create a particle physics simulation, using relevant literature to provide a more practical implementation of the mathematics written about
in particle physics simulation literature.

\subsection{Selected Particle Physics Simulations}

The following table summaries five Github projects that implement particle or sub-atomic particle simulations.
All of which are open-source and written in a language that are approachable to software engineers such as C++ or Python.

\begin{table}[ht]
    \centering
    \caption{Community Particle Simulation Reference Table}
    \label{tab:community-sims}
    \begin{tabularx}{\textwidth}{@{} L l l L @{}}
        \toprule
        Repository & Language & Visualisation & Accessibility \\
        \midrule
        Monte-Carlo Particle Simulator & C & None (CLI) & Very lightweight, however documentation requires compilation \\
        \addlinespace
        Charged-Subatomic-Particle-Interaction-Simulator & Python & Video & Physics equations buried in a single file, documentation non-existent \\
        \addlinespace
        Quantum Collider Sandbox & Python & 3D & Very comprehensive documentation \\
        \addlinespace
        APRIL \-- A Particle Runtime Interaction Library &  & &  \\
        \addlinespace
        OpenWave &  &  &  \\
        \bottomrule
    \end{tabularx}
\end{table}

\subsubsection*{Monte-Carlo Particle Simulator (gyaikhom)}

The simulation \bookcite{GyaikhomMontecarloparticlesimulator}{} implements a Monte-Carlo particle physics simulation model, specifically designed for high-energy physics.

\break{}

\textbf{Accessibility}

\begin{itemize}
    \item This project must be compiled, which may restrict some developers who are not familiar with compiling applications from source, from using it.
    \item The project is quite dated, with the last commit being from 12 years ago. This can make it challenging to compile due to the outdated CMake version it requires.
    \item CUDA is required to compile the project, which can prove to be a another barrier for a developer to use if they are unable to, or does not know, how to acquire it.
\end{itemize}

Due to this simulation outputting its result in the terminal, it may be hard to visualise the data that is being passed passed into it.
Given this, and its age, it can be confidently stated that this simulation may not be approachable for a developer to use.

\subsubsection*{Charged-Subatomic-Particle-Interaction-Simulator (RaresC21)}

The simulation \bookcite{cristianRaresC21ChargedSubatomicParticleInteractionSimulator2017}{} implements a simple model for displaying charged particle trajectories in a uniform magnetic field. It models protons and electrons but does \textbf{not} include neutrons or whole atoms/isotopes.
This simulation takes an interesting approach for its visualisation, opting for a \textit{.mp4} video rendered using \bookcite{FFmpeg}{} to display the results.
\\
\textbf{Accessibility}

\begin{itemize}
    \item This project requires Python to run.
    \item Given the project being written in Python, it is quite easy to modify the code.
    \item Documentation is non-existent. Requiring the user to reverse-engineer the code to understand how it works and modify it to thier preference.
\end{itemize}

Considering that this project only requires Python, a language found on most computers, this simulation is arguably quite accessible. However, the lack of documentation and explanatory code comments makes this project harder to modify to the users needs.
Furthermore, the burdersome act of having to run the python script, then opening a \textit{.mp4} file could prove unappealing to developers who would prefer a program where they can perform quick, iterative changes.

\subsubsection*{quantum-collider-sandbox (ml3m)}

The simulation \bookcite{mlemMl3mQuantumcollidersandbox2026}{} presents a 3D Sandbox scene which allows the user to view and manipulate various settings
\\
\textbf{Accessibility}

\begin{itemize}
    \item This project requires Python to run.
    \item Given the project being written in Python, it is quite easy to modify the code.
    \item Documentation is very comprehensive. A web page is provided which includes an installation guide and a usage guide.
    \item The simulation includes a user interface.
    \item Some editable fields have vague names.
\end{itemize}

This simulation is very accessible.
Due to this project only requiring Python to run, it is very easy to set up and run. Alongside this, the accessible comprehensive documentation helps to provide the user a very powerful sandbox.

Some of the editable fields in the interface suffer from vague names such as \textit{`B\_x'}, \textit{`B\_y'}, and \textit{`E\_x'}. These are explained as the magnetic and electromagnetic fields, this lack of immediate explanatory variable naming is very sparse.
It is not apparent if the user can 'spawn' in particles by themselves, which may hinder the accessibility when creating any unique situations that the user may desire.

\subsubsection*{APRIL \- A Particle Runtime Interaction Library (JKKDY)}

This framework \bookcite{deller-yeeJKKDYApril2026}{}, aims to provide the resources to create high-performance particle-based simulations. Whilst not strictly being a `simulation', it provides the tools
necessary to create them.
\\
\textbf{Accessibility}

\begin{itemize}
    \item This project is newer compared to the aforementioned projects. Last being updated in March 2026.
    \item Documentation is provided, including basic example projects and performance statistics.
    \item Being a framework requires more setup than other simulations, which is detrimental for the accessibility of this project.
    \item Since this is a C++ framework, some developers may struggle to use it given the complexity of the language.
\end{itemize}